---
title: "Traumatic Brain Injury"
author: "Jodie Lord"
date: "28/03/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, comment="")
```

### Task:

Use data from civilian and military populations to visualise type and frequency of traumatic brain injury, to help spread awareness for brain injury awareness month.

Full readme for task available within the **R for Data Science, TidyTuesday** github repository, available [here](https://github.com/rfordatascience/tidytuesday/tree/master/data/2020/2020-03-24)

_____________________________

#### Installing and loading libraries


```{r, message=FALSE, warning=FALSE}

library(tidyverse)
library(RColorBrewer)
#install.packages("pdftools")
library(patchwork)
library(cowplot)
library(ggtext)

```

_____________________________

#### Obtaining and inspecting data


```{r, message=FALSE, warning=FALSE}

tbi_age <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-03-24/tbi_age.csv')

tbi_age %>%
  head()  
  
tbi_age %>%  
  dim() %>% 

distinct(tbi_age, type)
distinct(tbi_age, age_group)


```

_____________________________

#### Preparing data

```{r, message=FALSE, warning=FALSE}

# renaming variables  
tbi_age$injury_mechanism<-recode(tbi_age$injury_mechanism, "Unintentional Falls"="Unintentional Fall", 
       "Unintentionally struck by or against an object"="Unintentionally Struck", 
       "Other unintentional injury, mechanism unspecified"="Mechanism Unspecified",
       "Motor Vehicle Crashes"="Motor Vehicle Crash",
       "Other or no mechanism specified"="Other/Unspecified")

# Creating factor levels for groups
tbi_age_levels<-c("0-4",
                  "5-14",
                  "15-24",
                  "25-34",
                  "35-44",
                  "45-54",
                  "55-64",
                  "65-74",
                  "75+")

tbi_type_levels<-c("Hospitalizations",
                  "Deaths",
                  "Emergency Department Visit")

tbi_unintended_levels<-c("Unintentional Fall",
                         "Unintentionally Struck",
                         "Mechanism Unspecified")


tbi_intended_levels<-c("Intentional self-harm",
                       "Motor Vehicle Crash",
                       "Assault",
                       "Other/Unspecified")

# filtering out null values and redundant age groups
tbi_age %<>%
  filter(!age_group %in% c("Total", "0-17")) %>%
  drop_na() %>%
  mutate(age_group=factor(age_group, levels=tbi_age_levels),
         type=factor(type, levels=tbi_type_levels)) 


# generating stratified dataframes
intentional<-tbi_age %>%
  filter(type %in% c("Hospitalizations", "Deaths"), !injury_mechanism %in% c("Unintentional Fall","Unintentionally Struck","Mechanism Unspecified")) %>%
  mutate(injury_mechanism=factor(injury_mechanism, levels=tbi_intended_levels))


unintentional_hosp<-tbi_age %>%
  filter(type == "Hospitalizations", injury_mechanism %in% c("Unintentional Fall","Unintentionally Struck","Mechanism Unspecified")) %>%
  mutate(injury_mechanism=factor(injury_mechanism, levels=tbi_unintended_levels))


unintentional_death<-tbi_age %>%
  filter(type == "Deaths", injury_mechanism %in% c("Unintentional Fall","Unintentionally Struck","Mechanism Unspecified")) %>%
  mutate(injury_mechanism=factor(injury_mechanism, levels=tbi_unintended_levels))


```

_____________________________

#### Plotting the data

```{r, message=FALSE, warning=FALSE}


# generating colour palettes
tbi_col<-c("#FF9900", "#336699", "#CCCCCC")

tbi_col_int<-c("#FFCC00", "#6699CC", "#336699", "#CCCCCC", "#CCCCCC", "#CCCCCC")


# wrapper <- function(x, ...)
# {
#   paste(strwrap(x, ...), collapse = "\n")
# }


# generating base plot
p1<-intentional %>% 
  ggplot(aes(x = age_group,
             y = rate_est, fill=injury_mechanism)) +
  scale_fill_manual(values=tbi_col_int) +
  facet_wrap(~type, nrow = 1, strip.position = "top") +
  geom_bar(stat="identity", width=0.7, colour="black") +
  labs(y = "Rate per 100,000", 
       x = "", 
       subtitle = "<span style='color:#6699CC'>**Motor vehicle crashes**</span> <span style='color:#666666'>pose the greatest known risk for TBI related</span> <span style='color:#6699CC'>**hospitaliztion**</span>.
<br><span style='color:#FFCC00'>**Intentional self-harm**</span><span style='color:#666666'> pose the greatest risk for TBI related </span><span style='color:#FFCC00'>**deaths**</span><span style='color:#666666'> in adult populations.</span>
<br>**Over 75s** <span style='color:#666666'>represent the </span>**highest risk age group overall** <span style='color:#666666'>for hospitalization or death due to TBI</span>") +
  ggtitle("<span style='color:#003399'>**CAUSES AND CONSEQUENCES OF TRAUMATIC BRAIN INJURY ACROSS AGES**
<br>----------------------------------------------------------------------------------------------------------------</span><br><br><span style='color:#666666'>Mechanisms of </span>**intentional injury**<span style='color:#666666'> posing greatest </span>**risk of hospitalization differ from**<span style='color:#666666'> <br> mechanisms posing greatest </span>**risk of death**") +
    theme(legend.title=element_blank(),
          panel.grid.major = element_blank(),  
          panel.grid.minor = element_blank(),
          panel.border = element_blank(),
          strip.text.x = element_text(size=12, face="bold"),
          plot.title=element_markdown(hjust=0),
          plot.subtitle=element_markdown(face = "italic", size=10))
  
# adding annotation text to deaths facet
dat_text <- data.frame(
  label = c("", "Deaths due to self-harm begin 
increasing from age 15."),
  type   = c("Hospitalizations", "Deaths"),
  injury_mechanism = c("Other/Unspecified", "Assault"),
  x     = c(2, 4),
  y     = c(80,60)
)

# adding arrow data to deaths facet
arrow_dat<-data.frame(type= c("Hospitalizations", "Deaths"),
  injury_mechanism = c("Other/Unspecified", "Assault"),
  x=c(0,4), xend=c(0.01,3), y=c(0,50), yend=c(0.01,20))

# adding annotation text and arrow to base plot
p1<-p1 + geom_text(
  data    = dat_text, size=2.5, fontface="italic", hjust=0.5,
  mapping = aes(x = x, y = y, label = label)
)+
  geom_curve(data=arrow_dat, 
             aes(x=x, y=y, xend=xend, yend=yend),
             inherit.aes=FALSE,arrow=arrow(length= unit(0.07, "npc")), linetype = 2, curvature = -0.2)



# generating second plot to visualise hospitalizations as a result of unintentional injuries
p2<-unintentional_hosp %>%
  ggplot(aes(x = age_group,
             y = rate_est,
             group=injury_mechanism,
             color=injury_mechanism)) +
    scale_color_manual(values=tbi_col) +
  geom_point() +
  geom_line(size=1) +
  ggtitle("**Hopitalization**<span style='color:#666666'> due to</span> **unintentional falls**<span style='color:#666666'> show small peak in infant age groups <br>and</span> **peak sharply**<span style='color:#666666'> again</span> **in the elderly**") +
  labs (y="Rate per 100,000",
        x= "Age group",
        subtitle = "<span style='color:#336699'>**Hospitalization**</span><span style='color:#666666'> due to</span><span style='color:#336699'> **other unintentional injuries**</span><span style='color:#666666'> remain comparably</span><span style='color:#336699'> **low**</span><span style='color:#666666'> across all age groups.</span><br>
        <span style='color:#FF9900'>**Deaths** </span><span style='color:#666666'>due to</span> <span style='color:#FF9900'>**unintentional falls**</span><span style='color:#666666'> (dashed orange line) are less likely in infants, but follow a similar <br></span><span style='color:#FF9900'>**upward risk**</span><span style='color:#666666'> in the </span><span style='color:#FF9900'>**elderly**</span><br>",
        caption="
        Graphic by: @jodielord5
        #TidyTuesday #BrainInjuryAwarenessMonth 
        Data source: www.cdc.gov/traumaticbraininjury/data | Year of Data: 2014 " ) +
    theme(legend.position="none",
        plot.title=element_markdown(),
        plot.subtitle=element_markdown(face = "italic", size=10),
        plot.caption = element_text(face="italic", color="grey"),
        panel.background = element_blank(),
        panel.grid.major = element_blank(),  
        panel.grid.minor = element_line(color="lightgrey"),
          panel.border = element_blank()) 

# creating new df subset confirming deaths caused by unintentional falls
fall_death<-unintentional_death%>%
  filter(injury_mechanism=="Unintentional Fall")

# adding in unintentional death info for more context
p3<-p2+geom_line(data=fall_death, size=0.5, linetype="dashed", color="orange", aes(x = age_group, y = rate_est)) +
  annotate("text",
           x=8.9,
           y=84,
           label=c("Unintentional fall 
deaths"),
           fontface="italic",
           size=2.8,
           color="#FF9900")


# generating line annotations
direct_labels<-unintentional_hosp %>%
  group_by(injury_mechanism) %>%
  summarize(y=max(rate_est)) 


direct_labels_axis<-axis_canvas(p3, axis="y")+
  geom_text(
    data=direct_labels,
    aes(y=y, label=injury_mechanism),
    x=-0.2,
    size=3.3,
    color=tbi_col,
    fontface="bold",
    hjust=0)


# inserting line annotations into visualisation
p_direct_labels<-insert_yaxis_grob(p3,direct_labels_axis)
#ggdraw(p_direct_labels)

# using patchwork to see both visualisations in single pane
final<-ggdraw(p1) /  ggdraw(p_direct_labels) 


# saving as image
ggsave("output_tbi.png",
  plot = final,
  width = 9,
  height = 12,
  units = c("in", "cm", "mm"),
  dpi = 300,
)


## END

```


